Application of a reaction-diffusion equation to the spread of the invasive ascidian *Botrylloides violaceus*

### 2024-07-23 PDEs/paper_outline.qmd
reorganizing files

# Abstract
Modeling is a critical tool for minimizing the harmful effects of invasive species on natural ecosystems. Despite the recent popularity of correlative environmental niche models, mechanistic reaction-diffusion models remain powerful tools for predicting and controlling the spread of invasive species. The colonial ascidian *Botrylloides violaceus* has rapidly spread across the Gulf of Maine, outcompeting and smothering native species and presenting a major nuisance for shellfish aquaculture.Here, we apply arelationship between the rate of spread of an invasive species and select biological parameters(namely, the population growth rate and diffusion coefficient) derived from the generalized reaction-diffusion equation. We validated the predicted rate of spread for *B. violaceus* against rates calculated from species occurrence data from the citizen science platform iNaturalist. The model predicted a mean speed of ~49 km/year, which was extremely close to the observed mean of ~40 km/year. Although simple reaction-diffusion models have limited biological realism, the accuracy of our results suggest that estimating spreading speed based on empirical data on local
diffusion potentials and population growth rates could be an underutilized approach in invasion ecology.

>>>>

# Abstract
Modeling is a critical tool for minimizing the harmful effects of invasive species on natural ecosystems. Despite the recent popularity of correlative environmental niche models, mechanistic reaction-diffusion models remain powerful tools for predicting and controlling the spread of invasive species. The colonial ascidian *Botrylloides violaceus* has rapidly spread across the Gulf of Maine, outcompeting and smothering native species and presenting a major nuisance for shellfish aquaculture. Here, we apply a relationship between the rate of spread of an invasive species and select biological parameters (namely, the population growth rate and diffusion coefficient) derived from the generalized reaction-diffusion equation. We validated the predicted rate of spread for *B. violaceus* against rates calculated from species occurrence data from the citizen science platform iNaturalist. The model predicted a mean speed of ~49 km/year, which was extremely close to the observed mean of ~40 km/year. Although simple reaction-diffusion models have limited biological realism, the accuracy of our results suggest that estimating spreading speed based on empirical data on local diffusion potentials and population growth rates could be an underutilized approach in invasion ecology.
